---
title: "OncoMix Vignette"
author: "Daniel Piqué"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
vignette: >
  %\VignetteIndexEntry{OncoMix Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```
 
## Introduction

Background: The advent of large, well-curated databases, such as the [genomic data commons](https://gdc.cancer.gov/), that contain RNA sequencing data from hundreds of patient tumors has made it possible to identify oncogene candidates based off of mRNA expression alone. Oncomix is the first method developed to identify oncogenes in a visually-interpretable manner from RNA-sequencing data in large cohorts of patients. 

OncoMix is an R package for identifying oncogene candidates based off of 2-component Gaussian mixture models. It estimates parameters using the expectation maximization procedure as implemented in the R package mclust. This tutorial will demonstrate how to identify oncogene candidates from a set of mRNA sequencing data. We start by loading the package:

```{r}
#devtools::install_github("dpique/oncomix")
library(oncomix)
```

We then pose the following question, the answer to which forms the basis for this method:

### What should the distribution of oncogene mRNA expression in a population of tumors and adjacent normal tissue ideally look like?

```{r}
library(ggplot2)
oncomix::oncoMixIdeal()
```

The x-axis represents mRNA expression values, with lower values toward the left and larger values (i.e. higher expression) toward the right. The y axis represents density. The teal curves represent the best-fitting Gaussian probability distribution (PD) over expression values from a single gene obtained from multiple tumor samples. The red-orange curves represent the PD over expression values from the same gene obtained from multiple adjacent normal tissue samples. This mixture model is applied once to the tumor data and again (separately) to the adjacent normal data, hence the 4 curves. 

The advantage of applying a 2-component mixture model is that we are able to capture biologically-relevant clusters of gene expression that may naturally exist in the data. Otherwise, we might represent our data with just a single curve sitting in the middle of what really are 2 distinct clusters. Visually, we see that for a theoretical oncogene, there is a *subgroup* of tumors that overexpresses this gene relative to normal tissue.

## Identify Oncogene Candidates 

Now, we will load at an example dataset that contains expression values for 4789 mRNA isoforms obtained from paired samples of breast tumor (`dfTumorIsof`) and adjacent normal(`dfNmlIsof`) breast tissue from 113 patients. We will fit the mixture model using the oncomix function `mixModelParams`, which takes dataframes that contain patients as rows and mRNA isoforms/genes as columns. The number of columns (genes) should be the same between both dataframes, though the number of rows can vary. 

```{r}
data(dfNmlIsof, dfTumorIsof, package="oncomix")

dim(dfNmlIsof)
dfNmlIsof[1:5, 1:5] #take a look at the matrix of mRNA expression data from adjacent normal samples

dim(dfTumorIsof)
dfTumorIsof[1:5, 1:5] #take a look at the matrix of mRNA expression data from tumors

mmParams = oncomix::mixModelParams(dfNmlIsof, dfTumorIsof) #fits the mixture models, will take a few minutes
head(mmParams)
```

The object returned by `mixModelParams` is a dataframe with rows corresponding to genes and 12 columns containing mixture model parameters. The rows are sorted according to the `score` column, with the first row containing the highest oncomix score (defined below). The meaning of the dataframe columns are described below:

  - `n.mu.1` = the mean ($\mu$) of the Gaussian curve with the smaller mean fit to the adjacent normal expression data 
  - `n.mu.2` = the mean ($\mu$) of the Gaussian curve with the larger mean fit to the adjacent normal expression data 
  - `n.var` = the variance ($\sigma$) of the two Gaussian curves fit to the adjacent normal expression data (fixed to be equal between the two curves)
  - `n.pi.1` = the proportion of adjacent normal samples assigned to the Gaussian curve with mean `n.mu.1`
  
  - `t.mu.1` = the mean ($\mu$) of the Gaussian curve with the smaller mean fit to the tumor expression data 
  - `t.mu.2` = the mean ($\mu$) of the Gaussian curve with the larger mean fit to the tumor expression data. Greater than 0 for sets of expression data with all values $\geq0$.
  - `t.var` = the variance ($\sigma$) of the two Gaussian curves fit to the tumor expression data (fixed to be equal between the two curves). Greater than 0.
  - `t.pi.1` = the proportion of tumor samples assigned to the Gaussian curve with mean `t.mu.1`. Between 0-1
  
  - `deltaMu2` = the difference between the means of the two curves between groups. `t.mu.2` - `n.mu.2`. May be negative or positive.  
  - `deltaMu1` = the difference between the means of the two curves between groups. `t.mu.1` - `n.mu.1`. May be negative or positive. 
  - `SI` = the selectivity index, or the proportion of adjacent normal samples with expression values less than the boundary defined by `t.mu.2` - `t.mu.1`. The selectivity index for the ith gene is computed as:
 
 $$SI_i  = \frac{1}{N}\sum_{j=1}^N \Bigg\{ \begin{array}{ll} 1,~ if~x_{ij} < \frac{\mu_{iLT}+\mu_{iHT}}{2}  \\ 
 0, ~ otherwise 
                  \end{array},
 $$

  - where $N$ is the number of adjacent normal samples, and $x_{ij}$ is the expression value of the ith gene in the jth adjacent normal sample. $\mu_{iHT}$ is the mean of higher/larger Gaussian from the ith gene in tumor samples, and $\mu_{iLT}$ is the mean of the smaller/lower Gaussian from the ith gene in the tumor samples.
  
  - `score` = The score for the $i^{th}$ gene is calculated as follows:
   $$score_i  = (\Delta\mu_{2i} - \Delta\mu_{1i}) - (var_{Ni} - var_{Ti}) ~~~~~, $$
 
  - where $\Delta\mu_{2i}$ and $\Delta\mu_{1i}$ are as described above (equivalent to `deltaMu2` and `deltaMu1`), $var_{Ni}$ is the common variance across the adjacent normal samples (equivalent to `n.var`), and $var_{Ti}$ is the common variance across the tumor samples (equivalent to `t.var`).

We can now use the `mmParams` dataframe to generate a variety of visualizations and to figure out which of our isoforms in our gene set is most similar in appearance to our idealized oncogene candidate.

For example, lets say that we wanted to select a subset of gene isoforms that most resembled the theoretically ideal oncogene. We can capture all of the genes meeting or exceeding specified thresholds using the `topGeneQuant` function. Here, we want to maximize `deltMu2`, minimize `deltMu1`, and maximize the `SI`.

```{r}
topGeneQuant = oncomix::topGeneQuants(mmParams, deltMu2Thresh = 99, deltMu1Thresh = 10, siThresh = .99)
print(topGeneQuant)
```

The results of the `topGeneQuants` function is to return a subsetted dataframe containing only those isoforms that met or exceeded the specified threshold. Here, there were `r nrow(topGeneQuant)` isoforms.

We can also select the top $N$ genes that most closely resemble the oncogene candidate by simply selecting the first $N$ rows from the `mmParams` object (e.g. `mmParams[1:N,]`). This is because the `mixModelParams` function returns a dataframe sorted by the score, with the 1st row containing the isoform with the highest score. There is an explanation of how the score is calculated above. 

```{r}
mmParams.top10 = mmParams[1:10,]
print(mmParams.top10)
```

## Visualize the output

### Visualize isoforms with a high SI & oncomix score

Now, we will visualize the distribution of gene expression values for a particular isoform. Specifically, we will create a overlapping histogram of an single isoform's expression values across both tumor (teal) and adjacent normal (red) samples with the best-fitting Gaussian curves superimposed. The isoform that we want to visualize here, uc002jxc.2, is among the 5 isoforms in the output from `topGeneQuant` function. It also ranks 7th out of 4789 isoforms in terms of its oncomix score, so it should have a distributional profile that is more similar to our theoretical oncogene that most other isoforms in this dataset.

```{r}
isof = "uc002jxc.2"
plotGeneHist(mmParams, dfNmlIsof, dfTumorIsof, isof)
```

Next, we will create a scatterplot with the axes corresponding to the differences between component means. Our oncogene candidates will be those genes that appear in the upper right quadrant of this scatterplot. The x axis corresponds to the difference between the means of the curves with the larger Gaussians (`deltaMu2`), and the y axis corresponds to the difference between the means of the curves with the smaller Gaussians (`deltaMu1`) between the two treatments.

Here, $\alpha$ (in the title) is a term that is present in the denominator of the value of the y-axis and functions as an automatic scaling parameter to set the range of the y-axis to be approximately equal to the range of the x-axis.

```{r, fig.width = 7, fig.height= 6.5}
scatterMixPlot(mmParams)
```

We would expect isoforms that maximize deltaMu1 and minimize deltaMu2 to be most visually similar to the theoretical oncogene candidate, and thus to be present within the uppre right quadrant of this histogram. However, due to the large variance displayed by some of these isoforms, not all isoforms in the upper right quadrant appear like the theoreticaly ideal oncogene. We developed an index, termed the selectivity index (SI), that helps highlight genes that follow our ideal profile. The SI ranges from 0 to 1, and genes with a larger selectivity index will follow more closely the ideal oncogene. Now, we will highlight the isoforms with a selectivity index greater than .99 using the `selIndThresh` argument to narrow our search.

```{r, fig.width = 7, fig.height= 6.5}
scatterMixPlot(mmParams, selIndThresh = .99)
```

We can also highlight where the top 10 isoforms with the highest score fall in the scatterplot.

```{r}
scatterMixPlot(mmParams = mmParams, gene_labels = rownames(mmParams.top10))
```

### Isoforms that map to oncogenes in an online database

We can check the distribution of the isoforms that map to known human oncogenes. To do this, we will use genes classified as oncogenes from the [ONGene database](http://ongene.bioinfo-minzhao.org/), which is first literature-curated database of oncogenes. This paper was published in 2017. 

We will then map the oncogenes in the database (which are in gene symbols) to ucsc isoforms, which is the gene format that we have in our original datasets. We'll show where the top 5 isoforms mapping to oncogenes in the ONGene database land on this scatterplot.

```{r}
#install.packages("RMySQL")
library(RMySQL)

#read in a table of known human oncogenes from the ONGene database
ongene = read.table("http://ongene.bioinfo-minzhao.org/ongene_human.txt", header = T, sep = "\t")

#send a sql query to UCSC to map the human oncogenes to ucsc isoform ids
ucsc_genome <- dbConnect(MySQL(), user="genome", 
                         host="genome-mysql.cse.ucsc.edu", db = 'hg19')
createGeneQuery = function(name){ #name is a character vector
  p1 = paste(name, collapse = '\',\'')
  p2 = paste('(\'',p1, '\')',sep ="")
  return(p2)
}
gene_q = createGeneQuery(ongene$OncogeneName)
query_res = dbGetQuery(ucsc_genome, paste0("SELECT kgID, geneSymbol FROM kgXref WHERE geneSymbol IN ", gene_q, " ;"))
dbDisconnect(ucsc_genome)

#Merge the query_res & mmParams dataframes
query_res$kgID.s = substr(query_res$kgID, 1,8)
mmParams$kgID.s = substr(rownames(mmParams), 1,8)
mmParams$kgID = rownames(mmParams)
mmParams.m = merge(mmParams, query_res, by = "kgID.s", all.x = T)
rownames(mmParams.m) = mmParams.m$kgID.x

# Show the top 5 isoforms with the highest score 
#in our dataset that map to known oncogenes
mmParams.m = mmParams.m[with(mmParams.m, order(-score)), ]
mmParams.m.s = subset(mmParams.m, !is.na(geneSymbol))[1:5,]
print(mmParams.m.s)

scatterMixPlot(mmParams = mmParams, gene_labels = rownames(mmParams.m.s))
```

If you are interested in a particular oncogene, then you can plug the name of that gene (as long as it is in your original dataset) into the `gene_labels` argument of the `scatterMixPlot` function, which will highlight those genes entered into the `gene_labels` argument on the scatterplot.

Note: Not all well-characterized oncogenes fall into or near the upper right quadrant of this scatterplot (explored more in the analysis below). This is expected because oncogenes can arise via a variety of mechanisms, such as via mutational activation (eg BRAF V600E), which may not be associated with increased expression of the gene. mRNA overexpresion is one of several ways that an oncogene can drive tumor behavior, and it is this class of oncogenes that our method seeks to detect.

### Distribution of the OncoMix score

Lastly, we will check the distribution of the OncoMix `score` across all gene isoforms. The histograms are superimposed and are colored by whether the isoform maps to a gene in the ONGene database (orange, n = `r table(is.na(mmParams.m$geneSymbol))[1]`) or not (purple, n = `r table(is.na(mmParams.m$geneSymbol))[2]`).

```{r}
library(ggplot2)
library(RColorBrewer)
col = brewer.pal(3, "Dark2")
ggplot(mmParams.m, aes(x = score, y = ..density.., fill=is.na(geneSymbol))) +
  geom_histogram(data=subset(mmParams.m, is.na(geneSymbol)), fill = col[2], alpha = 0.5)+ 
  geom_histogram(data=subset(mmParams.m, !is.na(geneSymbol)), fill = col[3], alpha = 0.5)+ theme_classic() + xlab("OncoMix Score")+ theme_classic() 
```

The distribution of the OncoMix scores looks equivalent between isoforms in the ONGene database versus those not in the ONGene database. However, the short right tail of this distribution, which contains isoforms with high scores, also preferentially consists of isoforms that map to genes in the ONGene database. In addition, the few isoforms that are not in the ONGene database and that have high scores may represent oncogenes that are yet to be discovered.

## Session Info

```{r}
sessionInfo()
```

Please email me at daniel.pique@med.einstein.yu.edu with any suggestions, questions, or comments. Thank you! 
