---
title: "OncoMix Vignette"
author: "Daniel Piqué"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
vignette: >
  %\VignetteIndexEntry{OncoMix Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```
 
## Introduction

OncoMix is an R package for identifying oncogene candidates based off of 2-component Gaussian mixture models. It estimates parameters using the expectation maximization procedure as implemented in the R package mclust. This tutorial will demonstrate how to identify oncogene candidates from a set of mRNA sequencing data.



The advent of large, well-curated databases, such as the [genomic data commons](https://gdc.cancer.gov/), that contain RNA sequencing data from hundreds of patient tumors has made it possible to identify oncogene candidates based off of mRNA expression alone. The method proposed here is the first developed to identify oncogenes from RNA-sequencing data in large cohorts of patients.

```{r}
#devtools::install_github("dpique/oncomix")
library(oncomix)
```

### What do we expect the distribution of an oncogene expression in a population to look like?

```{r}
library(ggplot2)
oncomix::oncoMixIdeal()

```

The teal curves represent the probability distribution (PD) over expression values from a single gene obtained from multiple tumors samples. The red-orange curves represent the PD over expression values from the same gene obtained from multiple adjacent normal tissue samples. Visually, we see that for a theoretical oncogene, there is a *subgroup* of tumors that overexpresses this gene relative to normal tissue.

## Identify Oncogene Candidates 

Now, we will load at an example dataset that contains expression values for 4789 mRNA  isoforms obtained from paired samples of breast tumor (`dfTumorIsof`) and adjacent normal(`dfNmlIsof`) breast tissue from 113 patients. We will fit the mixture model using the oncomix function `mixModelParams`, which takes dataframes that contain patients as rows and mRNA isoforms/genes as columns. The number of columns (genes) should be the same between both dataframes, though the number of rows can vary. 

```{r}
data(dfNmlIsof, dfTumorIsof, package="oncomix")

dim(dfNmlIsof)
dfNmlIsof[1:5, 1:5] #take a look at the matrix of mRNA expression data from adjacent normal samples

dim(dfTumorIsof)
dfTumorIsof[1:5, 1:5] #take a look at the matrix of mRNA expression data from tumors

mmParams = oncomix::mixModelParams(dfNmlIsof, dfTumorIsof) #fits the mixture models, will take a few mins
head(mmParams)
```

The object returned by `mixModelParams` is a matrix with rows corresponding to genes and 11 columns containing mixture model parameters. The meaning of the matrix columns are described below:

  - `n.mu.1` = the mean ($\mu$) of the Gaussian curve with the smaller mean fit to the adjacent normal expression data 
  - `n.mu.2` = the mean ($\mu$) of the Gaussian curve with the larger mean fit to the adjacent normal expression data 
  - `n.var` = the variance ($\sigma$) of the two Gaussian curves fit to the adjacent normal expression data (fixed to be equal between the two curves)
  - `n.pi.1` = the proportion of adjacent normal samples assigned to the Gaussian curve with mean `n.mu.1`
  
  - `t.mu.1` = the mean ($\mu$) of the Gaussian curve with the smaller mean fit to the tumor expression data 
  - `t.mu.2` = the mean ($\mu$) of the Gaussian curve with the larger mean fit to the tumor expression data. Greater than 0 for sets of expression data with all values $\geq0$.
  - `t.var` = the variance ($\sigma$) of the two Gaussian curves fit to the tumor expression data (fixed to be equal between the two curves). Greater than 0.
  - `t.pi.1` = the proportion of tumor samples assigned to the Gaussian curve with mean `t.mu.1`. Between 0-1
  
  - `deltaMu2` = the difference between the means of the two curves between groups. `t.mu.2` - `n.mu.2`. May be negative or positive.  
  - `deltaMu1` = the difference between the means of the two curves between groups. `t.mu.1` - `n.mu.1`. May be negative or positive. 
  - `SI` = the selectivity index, or the proportion of adjacent normal samples with expression values less than the boundary defined by `t.mu.2` - `t.mu.1`. The selectivity index for the ith gene is computed as:
 
 $$SI_i  = \frac{1}{N}\sum_{j=1}^N \Bigg\{ \begin{array}{ll} 1,~ if~x_{ij} < \frac{\mu_{iLT}+\mu_{iHT}}{2}  \\ 
 0, ~ otherwise 
                  \end{array},
 $$

  - where $N$ is the number of adjacent normal samples, and $x_{ij}$ is the expression value of the ith gene in the jth adjacent normal sample. $\mu_{iHT}$ is the mean of higher/larger Gaussian from the ith gene in tumor samples, and $\mu_{iLT}$ is the mean of the smaller/lower Gaussian from the ith gene in the tumor samples.
  

## Visualize the output

- We will create a scatterplot with the axes corresponding to the differences between component means. Our oncogene candidates will be those genes that appear in the upper right quadrant of this scatterplot. The x axis corresponds to the difference between the means of the curves with the larger Gaussians (`deltaMu2`), and the y axis corresponds to the difference between the means of the curves with the smaller Gaussians (`deltaMu1`) between the two treatments.

- Here, $\alpha$ is a term that is present in the denominator of the y-axis and functions as an automatic scaling parameter to set the range of the y- axis to be equal to the range of the x-axis.

```{r, fig.width = 7, fig.height= 6.5}
scatterMixPlot(mmParams)
```

We can see, however, that not all genes in the upper right quadrant appear like the theoreticaly ideal oncogene. We developed an index, termed the selectivity index, that helps highlight genes that follow this ideal profile. The SI ranges from 0 to 1, and genes with a larger selectivity index will follow more closely the ideal oncogene. Now, we will highlight the isoforms with a selectivity index greater than .99 to narrow our search.

```{r, fig.width = 7, fig.height= 6.5}
scatterMixPlot(mmParams, selIndThresh = .99)
```

We can also check the distribution of the selectivity index statistic across all gene isoforms to see whether the distribution is uniform or not.

```{r}
library(ggplot2)
qplot(mmParams[,"SI"]) + theme_classic() + xlab("Selectivity Index")

```

Lets say that we wanted to select a subset of gene isoforms that most resembled the theoretically ideal oncogene. We can capture all of the genes with prespecified thresholds using the `topGeneQuant` function. We can also select the top $N$ genes that most closely resemble the oncogene candidate using a custom score (`topGeneTable`). 

```{r}
mmParams.df = as.data.frame(mmParams)
topGeneQuant = oncomix::topGeneQuants(mmParams.df, deltMu2Thresh = 90, deltMu1Thresh = 10, siThresh = .99)
print(topGeneQuant)
topGeneTbl = oncomix::topGeneTable(mmParams.df, N = 10) #want the top 10 isoforms based on a custom score
print(topGeneTbl)

```


- Now, we will visualize the distribution of gene expression values for a particular isoform. Specifically, we will create a overlapping histogram of an single isoforms's expression values across both tumor (teal) and adjacent normal (red) samples with the best-fitting Gaussian curves superimposed.

```{r}
isof = "uc002jxc.2"
plotGeneHist(mmParams, dfNmlIsof, dfTumorIsof, isof)

```


```{r}
sessionInfo()
```

